Local PCA Shows How Population Structure Differs Along the Genome

### 2016-08-12 writeup/local_pca_paper.tex
.

\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
%\geometry{landscape}                		% Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex		
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{setspace}
\usepackage[square,sort,comma,numbers]{natbib} \bibpunct{(}{)}{;}{author-year}{}{,} 
\usepackage[hidelinks]{hyperref}

@@

\paragraph{\textit{Medicago truncatula}:}
Finally, we used whole-genome sequencing data from the \textit{Medicago truncatula} Hapmap Project \citep{tang2014improved},
which has 263 samples from 24 countries,
primarily distributed around the Mediterranean basin.
Each of the 8 chromosomes has 3--5 million SNPs;
PCA plots for these are shown in Figure \ref{fig:pca_medicago_allchr}.
we did not use the mitochondria or chloroplasts.

@@

To attempt to see how population structure varies without the effect of polymorphic inversions,
we then performed the same analyses after removing, for each chromosome arm,
the samples having the less frequent orientation of any inversion on that chromosome arm.
The result showed substantially less variation in MDS axes,
although some variation associated with inversion breakpoints was still visible,
as shown in Supplemental Figure \ref{sfig:mds_allchr_noinversions}.
For instance, the majority of the variation along 3L in Figure \ref{fig:mds_allchr}
is on the left end of the arm, dominated by two large peaks around the inversion breakpoints;
there is also a relatively small dip on the right end of the arm (near the centromere).
In contrast, Supplemental Figure \ref{sfig:mds_allchr_noinversions} shows that after removing polymorphic inversions,
remaining structure is dominated by the dip near the centromere.
The remaining variation seems to vary systematically along the chromosome,
following roughly similar patterns to that seen in recombination rate and diversity in \citet{langley2012genomic} and \citet{mackay2012drosophila}.

@@

>>>>

\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
%\geometry{landscape}                		% Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or epsÂ§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex		
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{setspace}
\usepackage[square,sort,comma,numbers]{natbib} \bibpunct{(}{)}{;}{author-year}{}{,} 
\usepackage[hidelinks]{hyperref}

@@

\paragraph{\textit{Medicago truncatula}:}
Finally, we used whole-genome sequencing data from the \textit{Medicago truncatula} Hapmap Project \citep{tang2014improved},
which has 263 samples from 24 countries,
primarily distributed around the Mediterranean basin.
Each of the 8 chromosomes has 3--5 million SNPs;
PCA plots for these are shown in Figure \ref{fig:pca_medicago_allchr}.
We did not use the mitochondria or chloroplasts.

@@

To see how population structure varies without the effect of polymorphic inversions,
we then performed the same analyses after removing, for each chromosome arm,
the samples having the less frequent orientation of any inversion on that chromosome arm.
The result showed substantially less variation in MDS axes,
although some variation associated with inversion breakpoints was still visible,
as shown in Supplemental Figure \ref{sfig:mds_allchr_noinversions}.
For instance, the majority of the variation along 3L in Figure \ref{fig:mds_allchr}
is on the left end of the arm, dominated by two large peaks around the inversion breakpoints;
there is also a relatively small dip on the right end of the arm (near the centromere).
In contrast, Supplemental Figure \ref{sfig:mds_allchr_noinversions} shows that after removing polymorphic inversions,
remaining structure is dominated by the dip near the centromere.
The remaining variation varies systematically along the chromosome,
following roughly similar patterns to that seen in recombination rate and diversity in \citet{langley2012genomic} and \citet{mackay2012drosophila}.
Indeed, correlations between the recombination rate \plr{XXX} near each window and the position on the first MDS coordinate are highly significant
(Figure \ref{fig:drosophila_recomb_rate}).
This is consistent with the hypothesis that variation in population structure
is due to selection, since the strength of linked selection increases with local gene density, measured in units of recombination distance.
However, the number of genes -- measured as the number of transcription start and end sites within each window (obtained from the RefGene table, UCSC genome browser) --
was more weakly correlated with MDS coordinate (Supplemental Figures \ref{sfig:drosophila_gene_density_plots} and \ref{sfig:drosophila_gene_density_correlations}).

@@

\begin{figure}
    \begin{center}
        \includegraphics{MDS_and_gene_count_allchr_drosophila_win104_try}
    \end{center}
    \caption{
        Gene number and population structure for \textit{Drosophila melanogaster}:
        The first MDS coordinate along the genome (black, above) is as shown in Supplemental Figure \ref{sfig:mds_allchr_noinversions};
        below is the number of genes' transcription start and end sites within each window, divided by two.
        Transcription start and end sites were obtained from the RefGene table from the UCSC browser, assembly dm3, on August 11, 2016.
        The extremely dense cluster of histone genes on chromosome arm 2L is excluded.
        \label{sfig:drosophila_gene_density_plots}
    }
\end{figure}

\begin{figure}
    \begin{center}
        \includegraphics{MDS_against_gene_count_allchr_drosophila_win104}
    \end{center}
    \caption{
        Gene number and population structure for \textit{Drosophila melanogaster}:
        this shows the first MDS coordinate and gene number, as in Supplementary Figure \ref{sfig:drosophila_gene_density_plots},
        against each other.
        The correlation coefficient and $p$-values for each linear regression are as follows:
        \plr{FILL THIS IN}.
% >summary of the linear regression of MDS against gene count for 2L
% 
% Call:
% lm(formula = coordinate1_2L[-k] ~ gene_count1[-k])
% 
% Residuals:
%      Min       1Q   Median       3Q      Max 
% -0.40257 -0.15326  0.04554  0.15037  0.32149 
% 
% Coefficients:
%                   Estimate Std. Error t value Pr(>|t|)
% (Intercept)      0.0087783  0.0182530   0.481    0.631
% gene_count1[-k] -0.0003592  0.0006379  -0.563    0.574
% 
% Residual standard error: 0.1851 on 273 degrees of freedom
% Multiple R-squared:  0.00116,	Adjusted R-squared:  -0.002498 
% F-statistic: 0.3172 on 1 and 273 DF,  p-value: 0.5738
% 
% 
% 
% > summary of the linear regression of MDS against gene count for 2R
% 
% Call:
% lm(formula = coordinate1_2R ~ gene_count2)
% 
% Residuals:
%      Min       1Q   Median       3Q      Max 
% -0.30622 -0.08482  0.02146  0.08438  0.30339 
% 
% Coefficients:
%               Estimate Std. Error t value Pr(>|t|)    
% (Intercept)  0.0846724  0.0163485   5.179 5.08e-07 ***
% gene_count2 -0.0028749  0.0004704  -6.112 4.52e-09 ***
% ---
% Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
% 
% Residual standard error: 0.1285 on 217 degrees of freedom
% Multiple R-squared:  0.1469,	Adjusted R-squared:  0.1429 
% F-statistic: 37.35 on 1 and 217 DF,  p-value: 4.516e-09
% 
% 
% 
% 
% > summary of the linear regression of MDS against gene count for 3L
% 
% Call:
% lm(formula = coordinate1_3L ~ gene_count3)
% 
% Residuals:
%      Min       1Q   Median       3Q      Max 
% -0.24754 -0.09927 -0.04406  0.09895  0.34383 
% 
% Coefficients:
%               Estimate Std. Error t value Pr(>|t|)    
% (Intercept) -0.0455675  0.0135826  -3.355  0.00091 ***
% gene_count3  0.0018695  0.0004338   4.309 2.31e-05 ***
% ---
% Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
% 
% Residual standard error: 0.1393 on 265 degrees of freedom
% Multiple R-squared:  0.06548,	Adjusted R-squared:  0.06196 
% F-statistic: 18.57 on 1 and 265 DF,  p-value: 2.312e-05
% 
% 
% 
% 
% 
% 
% > summary of the linear regression of MDS against gene count for 3R
% 
% Call:
% lm(formula = coordinate1_3R ~ gene_count4)
% 
% Residuals:
%     Min      1Q  Median      3Q     Max 
% -0.2883 -0.1316 -0.0424  0.0892  0.4433 
% 
% Coefficients:
%               Estimate Std. Error t value Pr(>|t|)    
% (Intercept) -0.0606512  0.0158644  -3.823 0.000161 ***
% gene_count4  0.0022703  0.0004608   4.927  1.4e-06 ***
% ---
% Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
% 
% Residual standard error: 0.1713 on 291 degrees of freedom
% Multiple R-squared:  0.077,	Adjusted R-squared:  0.07383 
% F-statistic: 24.28 on 1 and 291 DF,  p-value: 1.403e-06
% 
% 
% 
% 
% 
% > summary of the linear regression of MDS against gene count for X
% 
% Call:
% lm(formula = coordinate1_X ~ gene_count5)
% 
% Residuals:
%      Min       1Q   Median       3Q      Max 
% -0.46790 -0.09270  0.01922  0.11102  0.35063 
% 
% Coefficients:
%               Estimate Std. Error t value Pr(>|t|)    
% (Intercept)  0.0411519  0.0139012   2.960 0.003331 ** 
% gene_count5 -0.0020778  0.0005487  -3.787 0.000186 ***
% ---
% Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
% 
% Residual standard error: 0.1471 on 286 degrees of freedom
% Multiple R-squared:  0.04774,	Adjusted R-squared:  0.04441 
% F-statistic: 14.34 on 1 and 286 DF,  p-value: 0.0001862

        \label{sfig:drosophila_gene_density_correlations}
    }
\end{figure}
